---
title: |
   | UK Macro:
   | A view from financial markets 
author: Andrew Benito
date: 'January 29, 2025'
#date: '`r format(Sys.Date(), "%B %d, %Y")`'
institute: "Presentation at King's College London"
date-format: long
execute: 
  echo: false
  output: asis
format: 
  beamer: 
    navigation: horizontal
    header-includes: |
       \setbeamertemplate{navigation symbols}{}
       \setbeamertemplate{footline}[page number]
---

```{r, include=FALSE, echo=FALSE}
# Packages----
rm(list = ls())
lapply(c('knitr', 'here', 'tidyverse', 'openxlsx', 'tsbox', 'Rblpapi', 'data.table', 'ggtext',
         'MacrobondAPI', 'ggsci', 'patchwork', 'ggbeeswarm', 'sysfonts', 'readxl', 'gt', 'gtExtras',
         'kableExtra', 'showtext', 'hrbrthemes', 'janitor', 'ggrepel',
         'hablar', 'Rblpapi'),
       require, character.only = TRUE)
knitr::opts_knit$set(root.dir = here::here())
import_roboto_condensed()
blpConnect() 
options(warn = -1)

# Source Functions
source(here('functions','chartFormat.R'))
source(here('functions','functions.R'))

# Options----
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	fig.width=20, fig.height=10,
	dev = "cairo_pdf"
)

```

# Key Questions

- How is the UK economy **evolving**?    
    
    
- How is the UK's **policy mix** evolving?  
   
   
- Where might we be going badly **wrong**?  



# BoE Policy Settings

```{r, include=TRUE}
source(here('scripts','01_BoE_settings.R'))
p1
```


- Current rate-cutting cycle follows MPC's sharpest rate-hiking cycle.  
- QT has reversed add'l Gilts held from the GBP150bn QE decision [Nov 2020].  

# Evolving forms of Monetary Policy

```{r}
source(here('scripts','02_mon_pol_shocks.R'))
p1
```
- Managing market expectations - medium-term path for monetary policy has a stronger and more prolonged effect on inflation.  


# Evolving Market expectations

```{r, include=TRUE}
source(here('scripts','03_OIS_curves.R'))
ois1 

```
- Hard-to-learn features, eg productivity growth and r-star.  

# Zooming in... the past year

```{r, include=TRUE}
ois2 

```

- Expected rate cuts have moderated, coinciding with US election, UK fiscal news and upward revisions to UK infl outlook.


# An Evolving Inflation Outlook?

```{r, warning=FALSE}
source(here('scripts','04_BoE_forecasts.R'))
p1+p2

```
- BoE's Fan charts say little about the scale, persistence and source of macro news -- about 'how is the economy evolving?'

# Macro news: scale, persistence of source of news

```{r}
infl.forecasts

```
- Policy framework broadly robust to its largest inflation shock.

# More recent disinflation news

```{r}
infl.forecasts.latest

```

- Some of the faster-than-expected disinflation of the past year is judged to be temporary. Inflation forecasts now evolving higher.


# The pattern of BoE forecast errors

```{r}
p3+p4+p5

```
- Pattern of forecast errors points to repeated, negative supply shocks. (CPI: higher-than-expected; U: lower-than-expected; GDP: unbiased).



# Growth agenda -- Evolving expectations and out-turns

```{r, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
path <- here('scripts', '05b_tracking_consensus_gdp.R')
source(path)
plt

```

- Medium-term growth: innovation and reallocation effects dominate capital deepening.
- Is weak private investment a consequence rather than a cause of weak TFP and GDP growth? Easing planning rules can help.  


# Household real incomes

```{r, include=TRUE}
source(here('scripts','06_create_data_hh_income.R'))
rdi.plot

```
- The drag from elevated inflation and interest paid subsiding; household taxes a steady/persistent drag since 2023.


# Sterling asset prices


```{r, include=TRUE}
source(here('scripts','07_asset_prices.R'))
assets.plot

```

- Aim: distinguish sources of news: (i) UK macro (ii) BoE Monetary policy news (iii) US macro (iv) Fed policy and (v) 'Risk-on'.


# Decomposing changes in asset prices into macro 


```{r, include=TRUE, echo=FALSE, fig.align="center"}
here("plots", "1.GBP-OIS10y.png") %>%
    knitr::include_graphics(error = F)

```

- Results from a Bayesian VAR estimated under sign restrictions on daily, financial market data.
- OIS higher on US spillovers.


# Public spending shouldered by bond-holders + tax-payers
```{r, include=TRUE}
source(here('scripts','08_fiscal_main.R'))
spend.lt.plot

```
- The bond market recognises that politicians can be inclined towards making Bond-holders, not Tax-payers, shoulder Public spending.


# Debt Issuance approaching its limits?
```{r, include=TRUE}
source(here('scripts','09_DMO_Remit.R'))
dmo.remit
```


# Truss vs Reeves: financial markets' (15day) assessment

```{r}
source(here('scripts','10_truss_reeves.R'))
a.plot / b.plot

```
- **Truss**: a loss of (foreign) **investor confidence**, exacerbated by long-term investors (LDI). BoE interventions stabilised. 
- **Reeves**: a more standard injection of **aggregate Demand**, met with higher ST interest-rates. A subsequent deterioration in aggregate Supply or Risk premium?


# Bouts of risk premium

```{r}
correl.plot

```

- Instability reflects time-varying roles for different shocks.  
- 'normal' $\rho>0$ consistent with Demand news pushing Yields and Currency in same direction. Risk premium => opposing directions and $\rho<0$. 
- Reeves episode a much smaller and more temporary bout of risk premium than Truss mini-Budget. 


# Worsening fiscal arithmetic: a public debt ratchet

```{r, include=TRUE}
# Public sector debt
debt.lt.plot

```

-	One-off Supply shocks should be met with *higher* Public Debt, worked-off gradually.  
- Yet: r\uparrow, g\downarrow  imply a *lower* optimal Debt/GDP ratio.


# Fiscal choices based on Macro news

```{r}
source(here('scripts','08d_fiscal_choices.R'))
responses.scatter

```
- Spending some of proceeds of good news rather than 'paying-down' debt. Oct'24 Budget loose relative to that pattern.  
- Fiscal policy tightened at only 2 out of 21 fiscal events since 2015. Estimates of 'Headroom' small.  

# A plan to restore Fiscal Space? No.

```{r}
source(here('scripts','10_debt_stab_pri_bal.R'))
gt.dat

```

- A plan to stabilise Net Financial Liabilities does not (i) allow for more negative supply shocks or (ii) restore fiscal space.


# r\uparrow, g\downarrow

```{r}
# Evolving int rate costs and productivity
source(here('scripts','08e_historical_forecasts.R'))
interest.plot + prody.plot

```

- +50bp on market interest rates adds £10bn to public borrowing in 5Y.
- -1% from level of productivity adds c.£10bn to public borrowing in 5Y.


# Estimates of neutral rates drifting higher

```{r}
source(here('scripts','10_maps_rstar.R'))
maps.r.plot

```

- Risk of US fiscal policy pushing up on global neutral rates (r*).  
- If Output gap>0 and Disinflation stalling, 'r' is closer to 'r*' or neutral than thought.  

# g\downarrow and y* may be lower than thought

```{r}
source(here('scripts','11_create_data_output_gap.R'))
# prod.plot + og.plot
og.plot

```

# U*\uparrow 

```{r, include=TRUE, echo=FALSE, message=FALSE}
source(here('scripts','12_create_uk_beveridge.R'))
beveridge.plot

```

- Pre-Covid vs. latest U/V curve suggests an outward shift in U*. Sticky wage growth is another symptom. 

# Rising unemployment

```{r, include=TRUE}
dv_u + rec_ds

```
- Risk: U*\uparrow  is mis-interpreted as a cyclical effect.
- Would imply some additional inflationary pressure, keeping 'r' higher than otherwise.


# What have we learned?

- **Learning** about how the economy is evolving, while still **reassessing the starting point** of what we thought we cld take as given.   

- **Bayesians**: learning is continuous. But might we have to 'unlearn' something, previously taken as given? Imagine: jumping to a different branch of a decision tree.

- **Policy**: the misdiagnosed 'starting point' origin of large policy mistakes. In **Markets**: opportunities from these discontinuities (a la *Big Short*). 

- *UK Examples*: (i) BoE underestimating a positive output gap, post-Covid in 2022, with 'incrementalism' since then. (ii) BoE and markets assuming unrealistic fiscal tightening in 2024H1. Forward curves drift higher since then.


# Conclusions

**How is the UK economy evolving?** Disinflation that still depends on a restrictive monetary policy stance. Supply-side constrains GDP growth.    

**How is the UK policy mix evolving?** Withdrawal of both monetary restriction and fiscal support slower than expected a year ago. Fiscal policy as a driver of Monetary policy outlook, as well as Global news.

**Where might we be going badly wrong?**  
- Both Debt sustainability and Inflation persistence are sensitive to over-stating supply capacity and its trend.  
- Added risk that higher U is mis-diagnosed as slack (U-U\*) when it owes to higher structural U*. Bouts of risk premium.  
- Cyclical: Consumers may discount fiscal consolidation, while mon pol eases (Hawkish/Infl). Or, transmission of past monetary tightening delayed (Dovish/Growth at risk).

