---
title: |
   | BoE Watchers:
   | On monetary and fiscal policy interactions 
author: Andrew Benito
date: 'May 12, 2025'
#date: '`r format(Sys.Date(), "%B %d, %Y")`'
institute: "Eisler Capital"
date-format: long
execute: 
  echo: false
  output: asis
format: 
  beamer: 
    navigation: horizontal
    header-includes: |
       \setbeamertemplate{navigation symbols}{}
       \setbeamertemplate{footline}[page number]
---

```{r, include=FALSE, echo=FALSE}
# Packages----
rm(list = ls())
lapply(c('knitr', 'here', 'tidyverse', 'openxlsx', 'tsbox', 'Rblpapi', 'data.table', 'ggtext',
         'MacrobondAPI', 'ggsci', 'patchwork', 'ggbeeswarm', 'sysfonts', 'readxl', 'gt', 'gtExtras',
         'kableExtra', 'showtext', 'hrbrthemes', 'janitor', 'ggrepel', 'glue',
         'hablar', 'Rblpapi'),
       require, character.only = TRUE)
knitr::opts_knit$set(root.dir = here::here())
import_roboto_condensed()
blpConnect() 
options(warn = -1)

# Source Functions
source(here('functions','chartFormat.R'))
source(here('functions','functions.R'))

# Options----
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	fig.width=20, fig.height=10,
	dev = "cairo_pdf"
)

```


# Main messages

- **Markets**: risk premium episodes suggest the UK's fiscal buffer is uncomfortably small.
- **Macro**: (i) a need to restore fiscal space (ii) the UK output gap is still positive (iii) recent falls in Employment partly come from the Supply-side. 

MPC again risks having under-estimated 'starting point' Output gap and role of Supply in recent economic weakness.

These point to an improved **policy mix** that would incorporate add'l **fiscal consolidation**. Whether **monetary policy** eases by more depends on where that would leave the Output gap overall, Inflation persistence and the incidence of Risk premium episodes. 
 

# A public debt ratchet and an expanded BoE balance sheet

```{r, include=TRUE}
# Public sector debt
source(here('scripts','08_fiscal_main.R'))
source(here('scripts','01_BoE_settings.R'))

debt.lt.plot1950 + p1

```

- Debt to GDP triple its average in MPC's first decade.  
- QE shortened the effective maturity of Public liabilities.  

# r\uparrow, g\downarrow

```{r}
# Evolving int rate costs and productivity
source(here('scripts','08e_historical_forecasts.R'))
interest.plot + prody.plot

```

- +50bp on market interest rates now adds £10bn to public borrowing in 5Y.


# Revisions in market expectations of 'r'   

```{r, include=TRUE}
source(here('scripts','03_OIS_curves.R'))
ois1 

```
- Hard-to-learn features, eg r-star and trend productivity growth (Leland _et al_, 2024).  


# Stabilising debt, then restoring Fiscal Space?

```{r}
source(here('scripts','10_debt_stab_pri_bal.R'))
gt.dat

```


# Estimates of neutral rates drifting higher

```{r}
source(here('scripts','10_maps_rstar.R'))
maps.r.plot
ggsave(here::here('output', 'maps-r-star.png'), plot = maps.r.plot, width = 9, height = 5, dpi = 110, units = "in", device='png')

```

- If Output gap>0 (and Disinflation stalling), 'r' is closer to 'r*' than previously thought. 


# y* may be lower than thought, incl on U*\uparrow

```{r, include=TRUE, echo=FALSE, message=FALSE}
source(here('scripts','11_create_data_output_gap.R'))
source(here('scripts','12_create_uk_beveridge.R'))
og.plot + beveridge.plot

```

- Latest U/V curve suggests an outward shift in U* vs. pre-Covid. Sticky wage growth another symptom.


# The pattern of BoE forecast errors

```{r}
source(here('scripts','04_BoE_forecasts.R'))
p3+p4+p5

```
- Pattern points to repeated, negative supply shocks under-estimated by BoE. (CPI: higher-than-expected; U: lower-than-expected; GDP: unbiased).


# Bouts of Sterling risk premium

```{r}
source(here('scripts','07_asset_prices.R'))
correl.plot

```

- Instability reflects time-varying roles for different shocks.  
- 'normal' $\rho>0$ consistent with Demand news pushing Yields and Currency in same direction. Risk premium news => opposing directions and $\rho<0$. 
- Cx Reeves episode a much smaller and more temporary bout of risk premium than Truss mini-Budget. 


# Some quantification of Risk premium vs. Demand effects

<!-- ```{r} -->
<!-- 3-variable SVAR -->

<!-- ``` -->


# Risk premium complicates predictable MPC Comms 

```{r}
source(here('scripts','10_truss_january.R'))
a.plot + b.plot

```


# Fiscal reactions to macro news. More fiscal consolidation?

```{r}
source(here('scripts','08d_fiscal_choices.R'))
responses.scatter

```
- Fiscal policy tightened at 2 out of 21 fiscal events since 2015.  
- A quicker pace of consolidation to recognise a still positive Output Gap and provide a larger buffer against risk premium.



# How MPC interprets company responses to higher payroll taxes. 

```{r}
# Chart of DMP survey
source(here('scripts','12_DMP_response_taxes.R'))
responses.bar

```

- Employment cuts owe to weaker Supply-side, as much as aggregate Demand, and $U*\uparrow$.  


# Conclusions

- **Markets**: risk premium episodes suggest the UK's fiscal buffer is uncomfortably small.
- **Macro**: (i) a need to restore fiscal space (ii) the UK Output gap is still positive (iii) recent falls in Employment partly come from the Supply-side. 

MPC again risks having under-estimated 'starting point' Output gap and role of Supply in recent economic weakness.

These point to an improved **policy mix** that would incorporate add'l **fiscal consolidation**. Whether **monetary policy** eases by more depends on where that would leave the output gap overall, inflation persistence and the incidence of risk premium episodes.

