---
title: |
   | UK macro policy outlook - a case study
author: Andrew Benito
date: 'March 18, 2025'
#date: '`r format(Sys.Date(), "%B %d, %Y")`'
institute: 
#  - "Eisler Capital"
  - "Skema Business School"
date-format: long
execute: 
  echo: false
  output: asis
format: 
  beamer: 
    navigation: horizontal
    header-includes: |
       \setbeamertemplate{navigation symbols}{}
       \setbeamertemplate{footline}[page number]
       \usepackage{booktabs}
---

```{r, include=FALSE, echo=FALSE}
# Packages----
rm(list = ls())
lapply(c('knitr', 'here', 'tidyverse', 'openxlsx', 'tsbox', 'Rblpapi', 'data.table', 'ggtext',
         'MacrobondAPI', 'ggsci', 'patchwork', 'ggbeeswarm', 'sysfonts', 'readxl', 'gt', 'gtExtras',
         'kableExtra', 'showtext', 'hrbrthemes', 'janitor', 'ggrepel', 'glue', 'MASS',
         'hablar', 'Rblpapi', 'ggbeeswarm'),
       require, character.only = TRUE)
knitr::opts_knit$set(root.dir = here::here())
import_roboto_condensed()
blpConnect() 
options(warn = -1)

# Source Functions
source(here('functions','chartFormat.R'))
source(here('functions','functions.R'))

# Options----
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	fig.width=22, fig.height=10, # 20 x 10
	dev = "cairo_pdf"
)

```


# Main messages

- **Markets** \textcolor{red}{risk premium episodes} suggest the UK's fiscal buffer is too small.
- **Macro** (i) a need to restore fiscal space (ii) the UK output gap is still positive (iii) recent falls in employment partly come from the supply-side. 

A better \textcolor{blue}{policy mix} would incorporate add'l \textcolor{blue}{fiscal consolidation}. If spare capacity opens up, \textcolor{blue}{monetary policy} can ease by more. But the current policy mix, with too little near-term fiscal adjustment and too much monetary easing risks a positive output gap, inflation persistence and recurring risk premium episodes. 
 

# A public debt ratchet and an expanded BoE balance sheet

```{r, include=TRUE}
# Public sector debt
source(here('scripts','01_fiscal_main.R'))
source(here('scripts','01_BoE_settings.R'))

debt.lt.plot1970 + p1

```

- Debt to GDP three times its average in MPC's first decade.  
- QE shortened the effective maturity of public liabilities.  

# r\uparrow, g\downarrow. (r - g)\uparrow\uparrow

```{r}
# Evolving int rate costs and productivity
source(here('scripts','02_historical_forecasts.R'))
interest.plot + prody.plot

```

- +50bp on market interest rates now adds £7bn to public borrowing in 5Y.
- Fiscal and Monetary policies are not well co-ordinated when set with different views about future productivity growth. 

# Stabilising debt. Fiscal space?

```{r}
source(here('scripts','03_debt_stab_pri_bal.R'))
gt.dat

```

# Delayed timing of fiscal consolidation

```{r}
# created and calc in 08e_historical_forecasts.R
pb.plot + pb.plot.latest

```
- Even excl. 2020/21, 't+4' forecast error on the OBR's primary balance forecast averages \textcolor{red}{-2.7pp}.


# Market expectations of 'r'   

```{r, include=TRUE}
source(here('scripts','04_OIS_curves.R'))
ois1 

```
- Hard-to-learn medium-term features, eg r* and trend productivity growth lead to serially-correlated revisions (Leland _et al_, 2024).   


# Estimates of neutral rates drift higher

```{r}
source(here('scripts','05_maps_rstar.R'))
boxplot
ggsave(here::here('output', 'maps-r-star.png'), plot = maps.r.plot, width = 9, height = 5, dpi = 110, units = "in", device='png')

```

- If the output gap > 0, and disinflation is stalling, 'r' is closer to 'r*' than previously thought. 


# y* may be lower than thought, incl on U*\uparrow

```{r, include=TRUE, echo=FALSE, message=FALSE}
source(here('scripts','06_create_data_output_gap.R'))
source(here('scripts','06_create_uk_beveridge.R'))
og.plot + beveridge.plot

```

- Latest U/V curve suggests an outward shift in $U^*$ vs. pre-Covid. Sticky wage growth another symptom.


# The past pattern of BoE forecast errors

```{r}
source(here('scripts','07_BoE_forecasts.R'))
p3+p4+p5

```
- Pattern points to repeated, negative supply shocks under-estimated by BoE. (CPI: higher-than-expected; U: lower-than-expected; GDP: unbiased).


# Bouts of Sterling risk premium

```{r}
source(here('scripts','08_asset_prices.R'))
correl.plot

```

- Instability reflects changeable roles for different shocks.  
- \textcolor{blue}{Demand} news pushes Yields and Currency in same direction ($\rho>0$). \textcolor{red}{Risk premium} news pushes them in opposing directions ($\rho<0$).  
- Cx Reeves January episode a much smaller and more temporary bout of risk premium than Truss mini-Budget. 


<!-- # Quantifying the risk premium effect -->

<!-- <!-- ```{r} --> -->
<!-- <!-- 3-variable SVAR --> -->

<!-- <!-- ``` --> -->


<!-- - Regardless of the precise source of the shock, the fiscal buffer is too small.   -->


# Risk premiums complicate policy transmission + 'predictable' MPC Comms 

```{r}
source(here('scripts','09_truss_january.R'))
a.plot + b.plot

```

# Are Dovish MPC Comms becoming less credible?

```{r}
source(here('scripts','10_mpc_asset_reactions.R'))
mpc.20y + mpc.infl5

```

- Lower 1y OIS on MPC days as likely to push up on Gilt yields, and more so than before the mini-Budget.

# The fiscal reaction. Fiscal fatigue?   

```{r}
source(here('scripts','11_fiscal_choices.R'))
responses.scatter + bohn.plot

```
- Fiscal policy tightened at 2 of 21 fiscal events since 2015.  


# Conclusions

- **Markets** \textcolor{red}{risk premium episodes} suggest the UK's fiscal buffer is too small.
- **Macro** (i) a need to restore fiscal space (ii) the UK output gap is still positive (iii) recent falls in employment partly come from the supply-side. 

A better \textcolor{blue}{policy mix} would incorporate add'l \textcolor{blue}{fiscal consolidation}. If spare capacity opens up, \textcolor{blue}{monetary policy} can ease by more. But the current policy mix, with too little near-term fiscal adjustment and too much monetary easing risks a positive output gap, inflation persistence and recurring risk premium episodes. 